Retrofitting Effect Handlers onto OCaml

### 2020-11-12 retro-concurrency.tex
Updates

At the bottom of the stack, we have the |handler_info|, which contains the
pointer to parent fiber, and the closures for the value, exception and effect
cases. The closures are created by the translation described in
\S\ref{sec:static_semantics}; Multicore OCaml supports exception patterns in
addition to effect patterns in the same handler. This is followed by a block to
support DWARF unwinding, whose details we will return to later. Then, there is
a top-level handler that invokes the exception handler |clos_hexn|
after switching to the parent fiber. This is followed by the |pc| of the code
that invokes the value handler |clos_hval| after switching to the parent stack. The
stack is laid out such that when the computation in the handler returns,
control goes to the value handler.

@@

One challenge is to encode in DWARF directives the CFA when stacks are switched
at effect handlers, callbacks and external calls. When starting to run on a new
fiber, we insert DWARF bytecode to follow the |parent_fiber| pointer and
dereference the |saved_sp| to get the CFA (|saved_sp + 8|). During callbacks
into OCaml, we save the current system stack pointer in the |DWARF block| in
Figure~\ref{fig:fiber} to identify the CFA in the C stack. DWARF unwinding for
external calls is implemented by following a link to the current OCaml stack
pointer. With these changes, we get the same backtrace for the \texttt{meander}
program from \S\ref{sec:unwind}, modulo runtime system functions due to effect
handlers.

@@

\textbf{\textit{Web server.}} As one of our primary goals of adding effect handlers to
OCaml is to support direct-style concurrency, we implement a full-fledged
HTTP/1.1 web server using effect handlers (|MC|). We use effect handlers to
implement a user-level threading library, which spawns a thread per request. We
use \texttt{httpaf} library~\cite{httpaf} for HTTP handling, and
\texttt{libev}~\cite{libev} for the eventloop. We compare our implementation
against an Lwt version (|lwt|) which also uses \texttt{httpaf} and
\texttt{libev}. Unlike using effect handlers, Lwt version is written in monadic
style, and does not have a notion of thread per request. For comparison, we
also include a Go version (|go|) that uses \texttt{net/http}~\cite{nethttp}
package.

>>>>

At the bottom of the stack, we have the |handler_info|, which contains the
pointer to parent fiber, and the closures for the value, exception and effect
cases. The closures are created by the translation described in
\S\ref{sec:static_semantics}; Multicore OCaml supports exception patterns in
addition to effect patterns in the same handler. This is followed by a context
block needed for DWARF and GC bookkeeping with callbacks. Then, there is a
top-level handler that invokes the exception handler |clos_hexn| after
switching to the parent fiber. This is followed by the |pc| of the code that
invokes the value handler |clos_hval| after switching to the parent stack. The
stack is laid out such that when the computation in the handler returns,
control goes to the value handler.

@@

One challenge is to encode in DWARF directives the CFA when stacks are switched
at effect handlers, callbacks and external calls. When starting to run on a new
fiber, we insert DWARF bytecode to follow the |parent_fiber| pointer and
dereference the |saved_sp| to get the CFA (|saved_sp + 8|). During callbacks
into OCaml, we save the current system stack pointer in the |context block| in
Figure~\ref{fig:fiber} to identify the CFA in the C stack. DWARF unwinding for
external calls is implemented by following a link to the current OCaml stack
pointer. With these changes, we get the same backtrace for the \texttt{meander}
program from \S\ref{sec:unwind}, modulo runtime system functions due to effect
handlers.

@@

\textbf{\textit{Web server.}} As one of our primary goals of adding effect handlers to
OCaml is to support direct-style concurrency, we implement a full-fledged
HTTP/1.1 web server using effect handlers (|MC|). We use effect handlers to
implement a user-level threading library, which spawns a thread per request. We
use \texttt{httpaf} library~\cite{httpaf} for HTTP handling, and
\texttt{libev}~\cite{libev} for the eventloop. We compare our implementation
against an Lwt version (|lwt|) which also uses \texttt{httpaf} and
\texttt{libev}. Unlike using effect handlers, Lwt version is written in monadic
style, and does not have a notion of thread per request. For comparison, we
also include a Go version (|go|) that uses \texttt{net/http}~\cite{nethttp}
package. As both OCaml versions are single threaded, the Go benchmark is run with
\texttt{GOMAXPROCS=1}.
